---
title: "Class 12: Genome Informatics"
author: "Kris Price (PID: A17464127)"
format: pdf
---

> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

```{r}
library(tidyverse)

samples <- read.table("geno.txt")

table(samples$geno)

samples %>%
  group_by(geno) %>%
  summarize(
    median = median(exp)
  )
```

The A/A genotype has a sample size of 108 & median expression level of 31.25, A/G has a sample size of 233 & median expression level of 25.06, and G/G has a sample size of 121 & median expression level of 20.07.

> Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

```{r}
ggplot(samples, aes(x = geno, y = exp, fill = geno)) +
  geom_boxplot(notch = TRUE) +
  xlab("Genotype") +
  ylab("Expression Level")
```

Yes, it seems like the SNP decreases the expression level of the G allele of the ORMDL3 gene.