---
title: "Class 17: Obtaining and processing SRA datasets on AWS"
author: "Kris Price (PID: A17464127)"
format: pdf
---

## Downstream Analysis

```{r}
library(tximport)
library(rhdf5)

# setup the folder and filenames to read
folders <- dir(pattern="SRR21568*")
samples <- sub("_quant", "", folders)
files <- file.path( folders, "abundance.h5" )
names(files) <- samples

txi.kallisto <- tximport(files, type = "kallisto", txOut = TRUE)
```

```{r}
head(txi.kallisto$counts)
```

We can see how many transcripts we have for each sample:

```{r}
colSums(txi.kallisto$counts)
```

And how many transcripts are detected in at least one sample:

```{r}
sum(rowSums(txi.kallisto$counts) > 0)
```

Before subsequent analysis, we should filter out annotated transcripts with no reads:

```{r}
to.keep <- rowSums(txi.kallisto$counts) > 0
kset.nonzero <- txi.kallisto$counts[to.keep, ]
```

AND those with no change over the samples:

```{r}
keep2 <- apply(kset.nonzero, 1, sd) > 0
x <- kset.nonzero[keep2, ]
```

### Principal Component Analysis

We can now apply any exploratory analysis technique to this counts matrix! As an example, we will perform a PCA of the transcriptomic profiles of these samples:

```{r}
pca <- prcomp(t(x), scale = TRUE)

summary(pca)
```

Now we can use the first two PCs as a coordinate system for visualizing the summarized transcriptomic profiles of each sample:

```{r}
plot(pca$x[,1], pca$x[,2],
     col=c("blue","blue","red","red"),
     xlab="PC1", ylab="PC2", pch=16)
```

> Q. Use ggplot to make a similar figure of PC1 vs PC2 and a seperate figure PC1 vs PC3 and PC2 vs PC3.

PC1 vs. PC2:

```{r}
library(tidyverse)
library(ggrepel)

colData <- data.frame(condition = factor(rep(c("control", "treatment"), each = 2)))
rownames(colData) <- colnames(txi.kallisto$counts)

y <- as.data.frame(pca$x)

y$Condition <- as.factor(colData$condition)

ggplot(y, aes(PC1, PC2, col = Condition)) +
  geom_point() +
  geom_text_repel(label = rownames(y))
```

PC1 vs. PC3:

```{r}
ggplot(y, aes(PC1, PC3, col = Condition)) +
  geom_point() +
  geom_text_repel(label = rownames(y))
```


PC2 vs. PC3:

```{r}
ggplot(y, aes(PC2, PC3, col = Condition)) +
  geom_point() +
  geom_text_repel(label = rownames(y))
```


## Differential-Expression Analysis

We can use DESeq2 to complete differential-expression analysis:

```{r}
sampleTable <- data.frame(condition = factor(rep(c("control", "treatment"), each = 2)))
rownames(sampleTable) <- colnames(txi.kallisto$counts)
```

```{r, message = FALSE}
library(DESeq2)
```

```{r}
dds <- DESeqDataSetFromTximport(txi.kallisto,
                                sampleTable,
                                design = ~condition)

dds <- DESeq(dds)

res <- results(dds)
head(res)
```

These results can go on to be visualized and used for pathway analysis.

