---
title: "Class 7: Machine Learning 1"
author: "Kris Price (PID: A17464127)"
format: pdf
toc: true
---

## Background

Today we will begin our exploration of important machine learning methods, with a focus on **clustering** and **dimensionality reduction**.

To start testing these methods, let's make up some sample data to cluster where we know what the answer should be.

```{r}
hist(rnorm(3000, mean = 10))
```

> Q. Can you generate 30 numbers centered at +3 & 30 numbers centered at -3 taken at random from a normal distribution?

```{r}
tmp <- c(rnorm(30, 3),
         rnorm(30, -3))

x <- cbind(x = tmp, y = rev(tmp))
plot(x)
```

## K-means Clustering

The main function in "base R" for K-means clustering is called `kmeans()`, let's try it out.

```{r}
k <- kmeans(x, centers = 2)
k
```
> Q. What component of your k-means result object has the cluster centers?

```{r}
k$centers
```
> Q. What component of your k-means result object has the cluster size (i.e. how many points are in each cluster)?

```{r}
k$size
```

> Q. What component of your k-means result object has the cluster membership vector (i.e. the main clustering result: which points are in which cluster)?

```{r}
k$cluster
```

> Q. Plot the results of clustering (i.e. our data colored by the clustering result) along with the cluster centers.

```{r}
plot(x, col = k$cluster)
points(k$centers, col = "blue", pch = 15, cex = 2)
```

> Q. Can you run `kmeans()` again and cluster into 4 clusters and plot the results just like we did above with coloring by cluster & cluster centers shown in blue?

```{r}
k4 <- kmeans(x, 4)
plot(x, col = k4$cluster)
points(k4$centers, col = "blue", pch = 15, cex = 2)
```

> **Key-point:** `kmeans()` will always return the clustering that we ask for (this is the "K" or "centers" in K-means)!

## Hierarchical clustering

The main function for hierarchical clustering in base R is called `hclust()`. One of the main differences with respect to the `kmeans()` function is that you cannot just pass your input data directly to `hclust()` - it needs a "distance matrix" as input. We can get this from lots of places, including the `dist()` function.

```{r}
d <- dist(x)
hc <- hclust(d)
plot(hc)
```

We can "cut" the dendogram or "tree" at a given height to yield our clusters. For this, we use the function `cutree()`

```{r}
plot(hc)
abline(h = 10, col = "red")
groups <- cutree(hc, h = 10)
```
> Q. Plot our data `x` colored by the clustering result from `hclust()` and `cutree()`?

```{r}
plot(x, col = groups)
```

## Principal Component Analysis (PCA)

PCA is a popular dimensionality reduction technique that is widely used in bioinformatics.

### PCA of UK food data

Read data on food consumption in the UK:

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
x
```

It looks like the row names are not set properly. We can fix this:

```{r}
# rownames(x) <- x[, 1]
# x <- x[, -1]
```

A better way to do this is fix the row names assignment at import time:

```{r}
x <- read.csv(url, row.names = 1)
```

> Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

```{r}
dim(x)
```

There are 17 rows and 4 columns in x. You could use `dim(x)` or a combination of `nrow()` and `ncol()`.

> Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?

I prefer solving the 'row-names problem' while importing the csv file because if you re-run the `x <- x[, -1]` code, it will continuously delete the first column. If you re-run `x <- read.csv(url, row.names = 1)`, this issue will not happen.

> Q3: Changing what optional argument in the above barplot() function results in the following plot?

You can change the `beside` argument to FALSE instead of TRUE:

```{r}
barplot(as.matrix(x), beside=F, col=rainbow(nrow(x)))
```

> Q5: We can use the pairs() function to generate all pairwise plots for our countries. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?

```{r}
pairs(x, col=rainbow(nrow(x)), pch=16)
```
If a given point lies on the diagonal for a plot, that means that the two countries that are being plotted eat similar amounts of a food category. If a point is away from the diagonal, then the two countries eat more different amounts of a food category.

> Q6. Based on the pairs and heatmap figures, which countries cluster together and what does this suggest about their food consumption patterns? Can you easily tell what the main differences between N. Ireland and the other countries of the UK in terms of this data-set?

I think that England, Wales, and Scotland cluster together the most, which suggests that they have similar food consumption habits. I can only tell what the main differences between N. Ireland and the other countries are in the heatmap, but not very easily.

We can install the **pheatmap** package with the `install.packages()` command that we used previously. Remember that we always run this in the console and not a code chunk in our Quarto document.

```{r}
library(pheatmap)

pheatmap(as.matrix(x))
```

Of all these plots, only the `pairs()` plot was useful. This, however took a bit of work to interpret and will not scale when I am looking at much bigger datasets.

## PCA to the Rescue

The main function in "base R" for PCA is called `prcomp()`

```{r}
pca <- prcomp(t(x))
summary(pca)
```

> Q. How much variance is capturned in the first PC?

67.44% of variance.

> Q. How many PCs do I need to capture at least 90% of the total variance in the dataset?

Two PCs capture 96.5% of the total variance.

> Q. Plot our main PCA result. This is called different things depending on field of study (e.g. "PC plot", "ordination plot", "score plot", "PC1 vs. PC2 plot")

```{r}
attributes(pca)
```

To generate our PCA score plot, we want the `pca$x` component of the result object.

```{r}
pca$x
```

```{r}
my_cols <- c("orange", "red", "blue", "darkgreen")
plot(pca$x[, 1], pca$x[, 2], col = my_cols, pch = 16)
```

```{r}
library(ggplot2)

ggplot(pca$x) +
  aes(PC1, PC2) +
  geom_point(col = my_cols)
```

## Digging deeper (variable loadings)

How do the original variables (i.e. the 17 different foods) contribute to our new PCs?

```{r}
ggplot(pca$rotation) +
  aes(x = PC1, 
      y = reorder(rownames(pca$rotation), PC1)) +
  geom_col(fill = "steelblue") +
  xlab("PC1 Loading Score") +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))
```

> Q9: Generate a similar ‘loadings plot’ for PC2. What two food groups feature prominantely and what does PC2 maninly tell us about?

```{r}
ggplot(pca$rotation) +
  aes(x = PC2, 
      y = reorder(rownames(pca$rotation), PC2)) +
  geom_col(fill = "steelblue") +
  xlab("PC2 Loading Score") +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))
```

Soft drinks and fresh potatoes feature most prominently. PC2 tells us that these two groups greatly drive 29.05% of the total variation.